---
title: Statistical analysis of "Historically, the presidential inauguration has frequently been frigid in Washington",
author: Henry Linder (mhlinder@gmail.com)
date: January 13, 2017
output: html_document
---

In
"[Historically, the presidential inauguration has frequently been frigid in Washington](https://www.washingtonpost.com/news/capital-weather-gang/wp/2017/01/10/historically-the-presidential-inauguration-has-been-a-frigid-day-in-washington-d-c/)",
Justin Grieser considers historical weather data for presidential
election.

As a statistician, I think this type of small-data (twenty
observations!) is great, and this illustrates my own believe that a
"good" data analysis is one that situates data within a narrative and
doesn't mistake lots of numbers and statistical analysis for an
interesting story.

On the other hand, as a statistician, I can't resist the temptation to
put to work the tools of my field. Specifically, I was curious about
the author's assertion, "Warmer inaugurations typically favor incoming
Republican presidents." I immediately wondered: does the data really
support this conclusion?

I wasn't able to directly download the numbers from the article, but I
think ballpark guesses based on the bar charts are probably close
enough for this type of back-of-the-envelope analysis.


```{r echo = FALSE, message = FALSE}
library(magrittr)
library(dplyr)
```

A little setup in R gives us a nice little dataset with twenty observations.

```{r}
indata <-
    data.frame(year  = seq(2013, 1937, by = -4),
               party = factor(c("D", "D", "R", "R", "D", "D", "R",
                                "R", "R", "D", "R", "R", "D", "D",
                                "R", "R", "D", "D", "D", "D")),
               temp  = c(40, 38, 35, 36, 34, 40, 52, 7, 55, 28, 42,
                         35, 37, 23, 44, 48, 36, 34, 29, 33))
indata
```

First, there's no doubt that from a *numerical* perspective,
Republican inaugurations are warmer on average than Democratic ones.

```{r}
## Subset based on political party
temp_D <- indata %>% filter(party == "D") %>% use_series(temp)
temp_R <- indata %>% filter(party == "R") %>% use_series(temp)

mean(temp_D)
mean(temp_R)
```

But, is there statistical support for this conclusion?

I consider graphical plots of the data for each political party
separately.  The benchmark statistical technique to compare two
different populations is a t-test, and for those paying attention at
home, this type of analysis requires approximate normality of the two
populations.

As a quick reminder, a normal distribution looks like a bell-shaped curve:

```{r echo = FALSE}
plot(seq(-4, 4, length = 100) %>% dnorm, type = "l", xlab = "", ylab = "", main = "")
```

So, we can consider how well the temperature data matches this
shape. Typically, we use a histogram for this purpose, but because the
sample size is so small, this type of plot is difficult to read. A
kernel density plot is slightly easier to read.

```{r echo = FALSE}
par(mfrow = c(1, 2))
plot(density(temp_D), main = "Democratic temperatures",
     xlab = "Temp. (F)", ylab = "")
plot(density(temp_R), main = "Republican temperatures",
     xlab = "Temp. (F)", ylab = "")
```

The densities don't match the normal distribution perfectly, but they
look approximately symmetric, and the t-test is pretty robust to
deviations from normality, i.e., it works well even when the data
is imperfect.

This conclusion is confirmed by a normal probability plot. If the data
follow a line, the data is approximately normal. These data points do
fall on a line for the temperature data, so normality is a reasonable assumption.

```{r echo = FALSE}
par(mfrow = c(1, 2))
qqnorm(temp_D, main = "Temperatures at Democratic inaugurations: Normal probability plot")
qqline(temp_D)
qqnorm(temp_R, main = "Temperatures at Republican inaugurations: Normal probability plot")
qqline(temp_R)
```

This paves the way for us to consider a t-test of **whether the
average temperature at Democratic and Republican inaugurations is
statistically different**:

```{r}
model <- t.test(temp ~ party, data = indata)
```

For a quick-and-dirty assessment, we can consider a 95% confidence
interval for the true difference between the average temperatures for
inaugurations of each party.

```{r}
model$conf.int
```

**If the averages are equal, their difference should equal zero**, and
because the confidence interval includes zero, we conclude that
**there is no statistical support for the conclusion that the average
temperature on inauguration day is different for Republican and
Democratic presidents.**

Considering the confidence interval in this case is equivalent to
consider the (dreaded) p-value:

```{r}
model$p.value
```

The p-value is larger than 0.05 (much larger!), so we conclude that
the average temperatures are equal (and statistics professors
everywhere cringed a little bit).

This analysis suggests that, on average, there is *no* difference in
average temperature, which is unsurprising---it would be strange
indeed if the temperature somehow differed by the President's
political party.

It's also worth noting that because of the properties of the t-test,
the small sample size means the difference between the observed
averages would need to be very large to conclude that they are
different, and if we were really concerned about this, I would
recommend expand the scope of the analysis to include more historical
data. This would require a more complicated model than a simple
average, but might indicate additional structure in the data.

Finally, the author also cites National Weather Service forecasts for
January 20, and gives some really interesting historical
information. It is, of course, worth noting that is is well documented
that weather forecasts at a horizon of more than six days
are
[unrealiable](http://www.nws.noaa.gov/ost/climate/STIP/RServices/huq_032509.htm),
and are useful mainly as qualitative assessments, not exact point
estimates.

That said, six of the twenty elections Grieser considers saw
precipitation, which matches the NWS forecast of a "1 in 3 chance of
measurable precipitation in the nation's capital on Jan. 20."

